---
title: "GY7702 Assignment 1"
author: "Ben Coombs"
date: "14/11/2020"
output: pdf_document
---

# **R for Data Science - GY7702 Assignment 1**

## Loading libraries
Throughout this assignment, I will be using the libraries tidyverse and knitr.
Therefore it is a good idea to load them straight away.

```{r, message=FALSE}
library(tidyverse)
library(knitr)
```

## Question 1

### Q1.1

```{r}
#Create the vector of 25 numbers listed on the question paper
nums <- c(NA, 3, 4, 4, 5, 2, 4, NA, 6, 3, 5, 4, 0, 5, 7, 5, NA, 5, 2, 4, NA, 
          3, 3, 5, NA)
#Create a new vector of the same numbers, this time with missing values (NA 
#values omitted)
nums_new <- nums[!is.na(nums)]
#Check if all responses are strongly agree or strongly disagree
all(nums_new %>% is.element(c(1,7)))
```

A return of FALSE indicates that there were participants in the survey that did 
not either completely agree or completely disagree.

### Q1.2

```{r}
#Return the positions in the vector of elements that are participants responding 
#somehow agree or stronger (5 or greater).
which(nums_new >= 5)
```

## Question 2

### Q2.1

```{r}
# Load the installed library "palmerpenguins"
library(palmerpenguins)
```

### Q2.2

```{r}
#Create a table to show species, island, bill length and body mass of the 10 
#Gentoo penguins with the highest body mass
gentoo_penguins <- penguins %>%
  #Only want the table to show species, island, bill length and body mass
  select(species, island, bill_length_mm, body_mass_g) %>%
  #Only display rows of Gentoo penguins
  filter(species == "Gentoo") %>%
  #Arrange the rows by body mass largest to smallest
  arrange(-body_mass_g) %>%
  #Take the first 10 values, i.e. the 10 Gentoo penguins with the largest body 
  #mass
  slice_head(n = 10)

# Display the table
knitr::kable(gentoo_penguins)
```

### Q2.3

```{r, message=FALSE}
#Create a table of the average bill length per island ordered by average bill 
#length
island_bill_length <- penguins %>%
  #Only considering island and bill length
  select(island, bill_length_mm) %>%
  #Remove NA values to correctly summarise
  filter(!is.na(bill_length_mm)) %>%
  #Group rows by island
  group_by(island) %>%
  #Summarise the islands by the average bill length
  summarise(
    avg_bill_length_mm = mean(bill_length_mm)
  ) %>%
  #Arrange the table by highest to lowest average bill length
  arrange(-avg_bill_length_mm)

knitr::kable(island_bill_length)
```

### Q2.4

```{r, message=FALSE}
#Create a table to show the maximum, median and minimum proportion between bill 
#length and bill depth by species
bill_ratios <- penguins %>%
  #Only considering species, bill length and bill depth
  select(species, bill_length_mm, bill_depth_mm) %>%
  #Remove NA values to correctly summarise
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) %>%
  #Calculate and create a column for the bill ratio of each penguin
  mutate(
    bill_ratio = bill_length_mm / bill_depth_mm
  ) %>%
  #Group rows by species
  group_by(species) %>%
  #Summarise each species with their minimum, median, and maximum bill ratios
  summarise(
    min_proportion = min(bill_ratio),
    median_proportion = median(bill_ratio),
    max_proportion = max(bill_ratio)
  )

knitr::kable(bill_ratios)
```

## Question 3

### Q3.1

```{r}
#Load the csv data to a variable
covid_data <- read_csv("covid19_cases_20200301_20201017.csv")
```

### Q 3.2

```{r}
Ealing_complete_covid_data <- covid_data %>%
  #Create complete table with a row for each day and area
  tidyr::complete(specimen_date, area_name) %>%
  #Arrange the table by area name (A-Z) and specimen date (oldest to newest)
  arrange(area_name, specimen_date) %>%
  #Group by area name to avoid the risk of filling NA values in a later step 
  #with values from a different area
  group_by(area_name) %>%
  #Fill NA values of new and cumulative cases with the value of the previous 
  #date for that area, if possible
  tidyr::fill(newCasesBySpecimenDate, cumCasesBySpecimenDate) %>%
  #Replace any remaining NA values with 0
  tidyr::replace_na(list(newCasesBySpecimenDate = 0, 
                         cumCasesBySpecimenDate = 0)
                    ) %>%
  #Subset the assigned area
  filter(area_name == "Ealing") %>%
  #Ungroup to drop area_name column
  ungroup() %>%
  #Drop area_name column
  select(-area_name)
```

### Q 3.3

```{r}
#Copy of Ealing_covid_data as a new variable
Ealing_day_before <- Ealing_complete_covid_data

Ealing_covid_development <- Ealing_day_before %>%
  #Create new column reporting the day after the date reported in Specimen_date
  #column
  mutate(
    day_to_match = specimen_date + 1
  ) %>%
  #Drop the columns specimen_date and cumCasesBySpecimenDate
  select(-specimen_date, -cumCasesBySpecimenDate) %>%
  #Rename the newCasesBySpecimen date to newCases_day_before
  rename(newCases_day_before = newCasesBySpecimenDate) %>%
  #Join this table with Ealing_complete_covid_data where day_to_match is equal 
  #to specimen_date
  full_join(
    Ealing_complete_covid_data,
    by = c("day_to_match" = "specimen_date")
  ) %>%
  #Create new calculated column of new cases as a percentage of the number of 
  #cases of the previous day
  mutate(
    As_percentage_of_day_before = (
      newCasesBySpecimenDate/newCases_day_before
      )*100
  )

#Display the first few lines
Ealing_covid_development %>%
slice_head(n = 10) %>%
kable()
```

Days reporting 0 new cases lead to "Inf" being the value for 
As_percentage_of_day_before of the following day as R is being asked to divide 
by zero. Instead, I will replace these values with NA.

```{r}
#Replace "Inf" values in As_percentage_of_day_before with NA
Ealing_covid_development$As_percentage_of_day_before[
  is.infinite(Ealing_covid_development$As_percentage_of_day_before)
  ] <- NA

#Display the first few lines
Ealing_covid_development %>%
slice_head(n = 10) %>%
kable()
```
