---
title: "GY7702 Assignment 1"
author: "Ben Coombs"
date: "14/11/2020"
output: pdf_document
---

# **R for Data Science - GY7702 Assignment 1**

This document is my submission for Assignment 1 for R for Data Science (GY7702) at the University of Leicester. It was created using RMarkdown in RStudio and knitted to a pdf. Check out my **[Github repository](https://github.com/benc97/RfDS-coursework-1)** for any updates to this markdown and more information.

The questions were provided by Dr. Stefano De Sabbata.

## Loading libraries
Throughout this assignment, I will be making use of the libraries `tidyverse` and `knitr`. Therefore it is a good idea to load them straight away.

```{r message=FALSE}
library(tidyverse)
library(knitr)
```

## Question 1

### Create the vector of 25 numbers between 1 and 7 listed below, using the function c. Assume those values represent the answers to a survey question, following the scale presented further below.

```{r}
#Create the vector of 25 numbers listed on the question paper
nums <- c(NA, 3, 4, 4, 5, 2, 4, NA, 6, 3, 5, 4, 0, 5, 7, 5, NA, 5, 2, 4, NA, 3,
          3, 5, NA)
```

### Q1.1: write the code necessary to check whether all participants to the survey either completely disagree or completely agree, once the missing values are excluded.

```{r}
#Create a new vector of the same numbers, this time with missing values (NA 
#values) omitted
nums_new <- nums[!is.na(nums)]
#Check if all responses are strongly agree or strongly disagree
all(nums_new %>% is.element(c(1,7)))
```

A return of `FALSE` indicates that there were participants in the survey that did not either completely agree or completely disagree.

### Q1.2: write the code necessary to extract the indexes related to the participants in the survey who at least somehow agree or more.

```{r}
#Return the positions in the vector of elements that are participants responding 
#somehow agree or stronger (5 or greater)
which(nums_new >= 5)
```

## Question 2

### Q2.1: Install the library [`palmerpenguins`](https://allisonhorst.github.io/palmerpenguins/articles/intro.html)

```{r}
#The library "palmerpenguins" is already installed so I just need to load it
library(palmerpenguins)
```

### Q2.2: write the code necessary to create a table showing *species*, *island*, *bill length* and *body mass* of the 10 *Gentoo* penguins in the penguins table with the highest *body mass*.

```{r}
#Create a table to show species, island, bill length and body mass of the 10 
#Gentoo penguins with the highest body mass
gentoo_penguins <- penguins %>%
  #Only want the table to show species, island, bill length and body mass
  select(species, island, bill_length_mm, body_mass_g) %>%
  #Only display rows of Gentoo penguins
  filter(species == "Gentoo") %>%
  #Arrange the rows by body mass largest to smallest
  arrange(-body_mass_g) %>%
  #Take the first 10 values, i.e. the 10 Gentoo penguins with the largest body 
  #mass
  slice_head(n = 10)

#Display the table
kable(gentoo_penguins)
```

### Q2.3: write the code necessary to create a table showing the average *bill length* per *island*, ordered by average *bill length*.


```{r, message=FALSE}
#Create a table of the average bill length per island ordered by average bill 
#length
island_bill_length <- penguins %>%
  #Only considering island and bill length
  select(island, bill_length_mm) %>%
  #Remove NA values to correctly summarise
  filter(!is.na(bill_length_mm)) %>%
  #Group rows by island
  group_by(island) %>%
  #Summarise the islands by the average bill length
  summarise(
    avg_bill_length_mm = mean(bill_length_mm)
  ) %>%
  #Arrange the table by highest to lowest average bill length
  arrange(-avg_bill_length_mm)

#Display the table
kable(island_bill_length)
```

### Q2.4: write the code necessary to create a table showing the minimum, median and maximum proportion between *bill length* and *bill depth* by *species*.


```{r, message=FALSE}
#Create a table to show the maximum, median and minimum proportion between bill 
#length and bill depth by species
bill_ratios <- penguins %>%
  #Only considering species, bill length and bill depth
  select(species, bill_length_mm, bill_depth_mm) %>%
  #Remove any rows with NA values to correctly summarise
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) %>%
  #Calculate and create a column for the bill ratio of each penguin
  mutate(
    bill_ratio = bill_length_mm / bill_depth_mm
  ) %>%
  #Group rows by species
  group_by(species) %>%
  #Summarise each species with their minimum, median, and maximum bill ratios
  summarise(
    min_proportion = min(bill_ratio),
    median_proportion = median(bill_ratio),
    max_proportion = max(bill_ratio)
  )

#Display the table
kable(bill_ratios)
```

## Question 3

### Q3.1: write the code necessary to load the data from `covid19_cases_20200301_20201017.csv` to a variable named `covid_data`.

`covid19_cases_20200301_20201017.csv` contains information [**from the UK Government**](https://coronavirus.data.gov.uk/details/cases) on the number of new and cumulative cases of covid-19 from 1st March 2020 to 17th October 2020 in the UK.

```{r}
#Load the csv data to a variable
covid_data <- read_csv("covid19_cases_20200301_20201017.csv")
```
This message indicates that the CSV has been successfully loaded.

### Q3.2: write the code necessary to:
- Create a complete table, containing a row for each day and area.
- Replace `NA` values with the value available for the previous date.
- Replace the remaining `NA` values (those that don’t have a previous value available) with zero.
- Subset only the area assigned to your student ID in the table in the appendix.
- Drop the `area_name` column.
- Store the resulting table in a variable named `[area]_complete_covid_data` (substituting `[area]` in the name of the variable with the name of the area assigned to you).

```{r}
#Store this complete table that is being created into the variable
#named Ealing_complete_covid_data
Ealing_complete_covid_data <- covid_data %>%
  #Create complete table with a row for each day and area
  complete(specimen_date, area_name) %>%
  #Arrange the table by area name (A-Z) and specimen date (oldest to newest)
  arrange(area_name, specimen_date) %>%
  #Group by area name to avoid the risk of filling NA values in a later step 
  #with values from a different area
  group_by(area_name) %>%
  #Fill NA values of new and cumulative cases with the value of the previous 
  #date for that area, if possible
  fill(newCasesBySpecimenDate, cumCasesBySpecimenDate) %>%
  #Replace any remaining NA values with 0
  replace_na(list(newCasesBySpecimenDate = 0, 
                         cumCasesBySpecimenDate = 0)
                    ) %>%
  #Subset the assigned area
  filter(area_name == "Ealing") %>%
  #Ungroup to drop area_name column
  ungroup() %>%
  #Drop area_name column
  select(-area_name)

#Display the first five rows
Ealing_complete_covid_data %>%
  slice_head(n = 5) %>%
  kable()
```

### Q3.3: Starting from the table `[area]_complete_covid_data` created for *Question 3.2*:
- Create a copy of `[area]_complete_covid_data`, i.e., as another variable named `[area]_day_before`.
- Use the library [`lubridate`](https://lubridate.tidyverse.org/) to create a new column named `day_to_match` in the new table `[area]_day_before` that reports the day after the day reported in the column `specimen_date`, as a character value (e.g., if `specimen_date` is `"2020-10-10"`, `day_to_match` should be `"2020-10-11"`). This way, the cases registered on Oct 10th will be used as point of comparison for the cases registered on Oct 11th.
- Drop the `specimen_date` and `cumCasesBySpecimenDate` columns from the `[area]_day_before` table.
- Rename the `newCasesBySpecimenDate` column of the the `[area]_day_before` table to `newCases_day_before`.
- Join `[area]_day_before` with `[area]_complete_covid_data`, where the column `specimen_date` of `[area]_complete_covid_data` is equal to the column `day_to_match` of `[area]_day_before`.
- Calculate a new column in the joined table, containing the number of new cases as a percentage of the number of new cases of the day before.
- Store the resulting table in a variable named `[area]_covid_development`.


```{r}
#Create a copy of Ealing_covid_data as a new variable
Ealing_day_before <- Ealing_complete_covid_data %>%
  #Create new column reporting the day after the date reported in Specimen_date
  #column
  mutate(
    day_to_match = specimen_date + 1
  ) %>%
  #Drop the columns specimen_date and cumCasesBySpecimenDate
  select(-specimen_date, -cumCasesBySpecimenDate) %>%
  #Rename the newCasesBySpecimen date to newCases_day_before
  rename(newCases_day_before = newCasesBySpecimenDate)

#Store this table that is being created into the variable named 
#Ealing_covid_development
Ealing_covid_development <- Ealing_complete_covid_data %>%
  #Join this table with Ealing_complete_covid_data where day_to_match is equal 
  #to specimen_date
  full_join(
    Ealing_day_before,
    by = c("specimen_date" = "day_to_match")
  ) %>%
  #Create new calculated column of new cases as a percentage of the number of 
  #cases of the previous day
  mutate(
    As_percentage_of_day_before = (
      newCasesBySpecimenDate/newCases_day_before
      )*100
  )

#Display the first five lines of each table
Ealing_day_before %>%
  slice_head(n = 5)

Ealing_covid_development %>%
  #dropping cumCasesBySpecimenDate so that As_percentage_of_day_before can be
  #displayed
  select(-cumCasesBySpecimenDate) %>%
  slice_head(n = 5)
```

Days reporting 0 new cases lead to `Inf` being the value for `As_percentage_of_day_before` of the following day as R is being asked to divide by zero. Instead, I will replace these values with `NaN`.

```{r}
#Replace "Inf" values in As_percentage_of_day_before with NaN
Ealing_covid_development$As_percentage_of_day_before[
  is.infinite(Ealing_covid_development$As_percentage_of_day_before)
  ] <- NaN

#Display the first 5 lines
Ealing_covid_development %>%
  #dropping cumCasesBySpecimenDate so that As_percentage_of_day_before can be
  #displayed
  select(-cumCasesBySpecimenDate) %>%
  slice_head(n = 5)
```

### Q3.4: Write a short text (max 150 words) describing the development of new cases in the area over time, as evidenced by the table [area]_covid_development.

```{r, warning=FALSE}
ggplot(Ealing_covid_development, 
       aes(x = specimen_date, y = newCasesBySpecimenDate)
       ) +
  geom_line() +
  scale_x_date(date_breaks = "1 month", date_labels = "%d %b %y") +
  labs(x = "Date",
       y = "New cases",
       caption = "Figure 1: New cases between 01-03-2020 and 17-10-2020 for Ealing.")
```

Figure 1 shows us that cases started to increase rapidly from zero from early to mid-March, reaching on average around 30 new cases per day. When the initial lockdown was introduced, the rate of increase of new cases per day started to decline, though the variation in new cases per day data is quite large. New daily cases then began to follow a downward trend from mid-April, with the rate of decrease slowly tapering off until daily new cases sustained an average of just under 5 between the early June and August. From August, daily new cases began to rise exponentially, passing April levels by late September and reaching almost 90 new cases in one day in early October. A slight dip does appear in the very recent data, but fluctuation in the entire time series means that we can’t confidently interpret this as the start of a downward trend.

## Question 4

Download the file `lad19_population.csv` from BlackBoard, which contains a table detailing the population
per Local Authority District in the UK.

- Write the code necessary to load the population data into a variable and join this new information with the information from the file `covid19_cases_20200301_20201017.csv`.
- Analyse the data as you see fit. The end result should be one (max two) tables illustrating the spread of covid-19 in the area assigned to your student ID in the table in the appendix. For instance, the analysis could illustrate the development of cases over time, compared to the population, or it could illustrate a comparison with other areas in the region.
- Include a short text (max 250 words) providing a short description of the analysis and interpretation of the results.


```{r}
#Load the csv data to a variable
lad_pop <- read_csv("lad19_population.csv")
```
This message indicates that the CSV has been successfully loaded.

```{r}
#Join the local authority info with the covid_data
lad_covid <- lad_pop %>%
  full_join(
    covid_data,
    by = c("lad19_area_name" = "area_name")
  )
```

I want use the population data for Ealing from `lad_covid` to calculate a new column `Ealing_covid_development` that reports the number of new cases per 100,000 of the population for that day. This will allow for comparison with other areas.

```{r}
Ealing_covid_pop <- lad_covid %>%
  #Include only rows related to Ealing from lad_covid
  filter(lad19_area_name == "Ealing") %>%
  #Consider only the area_population and specimen_date column
  select(area_population, specimen_date) %>%
  #Join the area population details onto Ealing_covid_development from Q3.3
  full_join(
    Ealing_covid_development,
    by = c("specimen_date" = "specimen_date")
  ) %>%
  #Calculate new column reporting new daily cases per 100k
  mutate(
    new_cases_per_100k = (newCasesBySpecimenDate/area_population)*100000
  )

#Display first 5 rows of table to show new cases per 100,000 column has been
#correctly calculated.
Ealing_covid_pop %>%
  select(specimen_date, 
         area_population, 
         newCasesBySpecimenDate, 
         new_cases_per_100k
         ) %>%
  slice_head(n = 5) %>%
  kable()
```

Obviously, in order to compare with another area, I need to obtain these figures for at least one other area. I will create a set of three functions that will be nested like a set of Russian dolls so that only a single function needs to be called to get the equivalent of `Ealing_covid_pop` for any area.

```{r}
#First of all, I need the complete table containing a row for each day and area
complete_covid_data <- covid_data %>%
  #Create complete table with a row for each day and area
  complete(specimen_date, area_name) %>%
  #Arrange the table by area name (A-Z) and specimen date (oldest to newest)
  arrange(area_name, specimen_date) %>%
  #Group by area name to avoid the risk of filling NA values in a later step 
  #with values from a different area
  group_by(area_name) %>%
  #Fill NA values of new and cumulative cases with the value of the previous 
  #date for that area, if possible
  fill(newCasesBySpecimenDate, cumCasesBySpecimenDate) %>%
  ungroup() %>%
  #Replace any remaining NA values with 0
  replace_na(list(newCasesBySpecimenDate = 0, 
                         cumCasesBySpecimenDate = 0)
                    )

#Display first 5 rows of table
complete_covid_data %>%
  slice_head(n = 5) %>%
  kable()
```

Now I can use this table to create a function to get complete covid data of any area, and save it to the variable `Area_complete_covid_data`.

```{r}
Area_complete_covid_data <- function(area) {
  complete_covid_data %>%
  #Subset the assigned area
  filter(area_name == area) %>%
  #Drop area_name column
  select(-area_name)
}

#Display first 5 rows of table with Powys as an example
Area_complete_covid_data("Powys") %>%
  slice_head(n = 5) %>%
  kable()
```

Now I can use this function to create another function to return the equivalent of `Ealing_covid_development` for any given area, and save it to the variable `Area_covid_development`.

```{r}
Area_covid_development <- function(area) {
  covid_dev <- Area_complete_covid_data(area) %>%
  #Create new column reporting the day after the date reported in Specimen_date
  #column
  mutate(
    day_to_match = specimen_date + 1
  ) %>%
  #Drop the columns specimen_date and cumCasesBySpecimenDate
  select(-specimen_date, -cumCasesBySpecimenDate) %>%
  #Rename the newCasesBySpecimen date to newCases_day_before
  rename(newCases_day_before = newCasesBySpecimenDate) %>%
  #Join this table with Ealing_complete_covid_data where day_to_match is equal 
  #to specimen_date
  full_join(
    Area_complete_covid_data(area),
    by = c("day_to_match" = "specimen_date")
  ) %>%
  #Create new calculated column of new cases as a percentage of the number of 
  #cases of the previous day
  mutate(
    As_percentage_of_day_before = (
      newCasesBySpecimenDate/newCases_day_before
      )*100
  )
  
  #Replace "Inf" values in covid_dev with NaN
  covid_dev$As_percentage_of_day_before[
  is.infinite(covid_dev$As_percentage_of_day_before)
  ] <- NaN
  
  covid_dev
}

#Display first 5 rows of table with Powys as an example, dropping 
#cumCasesBySpecimenDate to show As_percentage_of_day_before
Area_covid_development("Powys") %>%
  select(-cumCasesBySpecimenDate) %>%
  slice_head(n = 5)
```

Now I can use `Area_covid_development` in a function to return the `Ealing_covid_pop` table for any area. I am going to use the function to get the daily new infection rate per 100,000 for the area Powys. 

```{r}
Area_covid_pop <- function(area) {
  covid_pop <- lad_covid %>%
    #Include only rows related to the input area
    filter(lad19_area_name == area) %>%
    #Take only area_population and specimen_date
    select(area_population, specimen_date) %>%
    #Join the area population details onto Ealing_covid_development from Q3.3
    full_join(
      Area_covid_development(area),
      by = c("specimen_date" = "day_to_match")
    ) %>%
    mutate(
      new_cases_per_100k = (newCasesBySpecimenDate/area_population)*100000
    )
}

#Display first 5 rows of table with Powys as an example, dropping 
#newCases_day_before, cumCasesBySpecimenDate and As_percentage_of_day_before 
#to display new_cases_per_100k
Area_covid_pop("Powys") %>%
  select(-newCases_day_before, 
         -cumCasesBySpecimenDate, 
         -As_percentage_of_day_before
         ) %>%
  slice_head(n = 5)
```

Now I can join `Ealing_covid_pop` to an identical table for Powys.
```{r}
Ealing_v_Powys <- Ealing_covid_pop %>%
  full_join(Area_covid_pop("Powys"),
            by = c("specimen_date" = "specimen_date")
  )

#Display first 5 rows of table
Ealing_v_Powys %>%
  slice_head(n = 5)
```

In this table, columns ending in `.x` are referring to Ealing, and columns ending in `.y` are referring to Powys, so I should rename them to be clear what they represent.

```{r, message=FALSE, warning=FALSE}
#Select specimen_date and the columns displaying new cases per 100,000 for each
#area.
Ealing_v_Powys_per_100k <- Ealing_v_Powys %>%
  select(specimen_date, new_cases_per_100k.x, new_cases_per_100k.y) %>%
  #Rename the columns to be clear what they represent
  rename(Ealing_new_cases_per_100k = new_cases_per_100k.x, 
         Powys_new_cases_per_100k = new_cases_per_100k.y
         )

#Display first 5 rows of table
Ealing_v_Powys_per_100k %>%
  slice_head(n = 5) %>%
  kable()

#Plot both new case rates per 100,000 people per day on the same chart
Ealing_v_Powys_per_100k %>%
  ggplot(aes(x = specimen_date)) +
  geom_smooth(aes(y = Ealing_new_cases_per_100k), colour = 'blue') +
  geom_smooth(aes(y = Powys_new_cases_per_100k), colour = 'red') +
  scale_x_date(date_breaks = "1 month", date_labels = "%d %b %y") +
  labs(x = "Date",
       y = "New cases per 100,000",
       caption = "Figure 2: Comparison of new cases per 100,000 between 01-03-2020 and 17-10-2020 for Ealing (blue) and Powys (red).")
```

With the local authority data providing population statistics for each area, I calculated the daily new COVID-19 cases for Ealing per 100,000 of the population. The data shows that it was a very small proportion of the population getting infected each day, though the most recent figures showed that despite lockdown, almost 1% of the population had since the beginning of March tested positive for COVID-19. I created a function to calculate new cases per 100,000 people per day for any area and compared the rate of new infections in Ealing to Powys. I thought they would be interesting to compare as Ealing is a densely populated urban area and Powys is a scarcely populated rural area. Figure 2 shows that both areas followed the same general trend, with cases rising from March into April, reaching a peak, decreasing during lockdown and levelling in the summer, only to begin to climb again from the beginning of August. Ealing’s initial peak occurred earlier than Powys’, but the rates were much higher, some days reaching above 12 per 100,000 whereas in Powys the number only once went above 10. Both areas saw similar low figures during late June and through July, but from August to early October, Ealing saw a much more drastic increase in new cases, some days recording over 25 new cases per 100,000. Although Powys’ cases are also rising, the difference between the two areas is largest in the most recent data.